### 2018-08-15 abstract.tex

Current interest in advanced nuclear energy and \gls{MSR} concepts has enhanced 
demand for tools to analyze these systems. This paper introduces a Python 
script, SaltProc, which simulates \gls{MSR} online reprocessing by modeling the 
changing isotopic composition of an irradiated fuel salt. SaltProc couples with 
the Monte Carlo code, SERPENT 2, for neutron transport and depletion 
calculations. SaltProc capabilities was illuminated for full-core depletion in 
the \gls{MSBR} which demonstrated that (1) equilibrium fuel composition could 
be achieved after 16 years and (2) the multiplication factor stabilizes after 6 
years of operation. Fuel salt irradiation with online reprocessing causes 
considerable neutron energy spectrum hardening; this spectral shift has a 
problematic impact on safety parameters (e.g., temperature reactivity feedback, 
reactivity control system worth). Volatile gases and noble metals removal has 
positive effect on the core performance approximately 7500 pcm over 10-years 
timeframe; rare earth elements removal has slower rate and less impact on 
reactivity (5500 pcm). Finally, the average $^{232}$Th feed rate throughout 60 
years of operation is 100 g/GWh$_e$ which is a good agreement with other recent 
research. Problematic effects of neutron energy spectrum hardening during 
\gls{MSBR} operation should be taken into account for neutronics, 
multi-physics, and fuel cycle performance analysis.

>>>>

Current interest in advanced nuclear energy and \gls{MSR} concepts has enhanced 
demand for tools to analyze these systems. This paper introduces a Python 
package, SaltProc, which simulates \gls{MSR} online reprocessing by modeling the 
changing isotopic composition of an irradiated fuel salt. SaltProc couples with 
the Monte Carlo neutron transport software, SERPENT 2, for neutron transport and depletion 
calculations. SaltProc capabilities was illuminated for full-core depletion in 
the \gls{MSBR} which demonstrated that (1) equilibrium fuel composition could 
be achieved after 16 years and (2) the multiplication factor stabilizes after 6 
years of operation. Fuel salt irradiation with online reprocessing causes 
considerable neutron energy spectrum hardening; this spectral shift has a 
problematic impact on safety parameters (e.g., temperature reactivity feedback, 
reactivity control system worth). Volatile gases and noble metals removal has 
positive effect on the core performance approximately 7500 pcm over 10-years 
timeframe; rare earth elements removal has slower rate and less impact on 
reactivity (5500 pcm). Finally, the average $^{232}$Th feed rate throughout 60 
years of operation is 100 g/GWh$_e$ which is a good agreement with other recent 
research. Problematic effects of neutron energy spectrum hardening during 
\gls{MSBR} operation should be taken into account for neutronics, 
multi-physics, and fuel cycle performance analysis.

### 2018-08-15 acks.tex

Kathryn D. Huff directed and 
supervised the work, conceived and designed the simulations, wrote the paper, 
contributed to the software product, and reviewed drafts of the paper.  Prof. 
Huff is supported by the Nuclear Regulatory Commission Faculty Development 
Program, the National Center for Supercomputing Applications, the NNSA Office 
of Defense Nuclear Nonproliferation R\&D through the Consortium for Verfication 
Technologies and the Consortium for Nonproliferation Enabling Capabilities, and 
the International Institute for Carbon Neutral Energy Research (WPI-I2CNER), 
sponsored by the Japanese Ministry of Education, Culture, Sports, Science and 
Technology.

>>>>

Kathryn D. Huff directed and 
supervised the work, conceived and designed the simulations, 
contributed to the software product, and reviewed drafts of the paper.  Prof. 
Huff is supported by the Nuclear Regulatory Commission Faculty Development 
Program, the National Center for Supercomputing Applications, the NNSA Office 
of Defense Nuclear Nonproliferation R\&D through the Consortium for Verfication 
Technologies and the Consortium for Nonproliferation Enabling Capabilities, and 
the International Institute for Carbon Neutral Energy Research (WPI-I2CNER), 
sponsored by the Japanese Ministry of Education, Culture, Sports, Science and 
Technology.

### 2018-08-15 conclusion.tex

\section{Conclusion}

Condensed copy-paste of the thesis conclusion.
\begin{itemize}
    \item Full-core model which is better the most exists. Importance of full-core approach for multi-region designs.
    \item K-eff dynamics and explanation of this dynamics.
    \item Spectral shift explanation.
    \item Why spectral shift causes safety parameters worsening and power profile changes.
    \item $^{232}$Th rate is in a good agreement with references.
\end{itemize}

\FloatBarrier

>>>>

\section{Discussion and conclusions}

This work introduces the open source \gls{MSR} simulation package SaltProc. 
SaltProc expands the capability of SERPENT 2, the continuous-energy Monte Carlo 
code to include online reprocessing in liquid-fueled \gls{MSR} operation 
\cite{andrei_rykhlevskii_arfc/saltproc:_2018}. Benefits of SaltProc include 
generic geometry modeling, multi-flow capabilities, time-dependent feed and 
removal rates, and the ability to specify removal efficiency. The main goal of 
this work has 
been to demonstrate the SaltProc capability to find the equilibrium fuel salt 
composition (where equilibrium is defined as when the number densities of major 
isotopes vary less than 1\% over several years). A secondary goal has been to 
compare predicted operational and safety parameters (e.g., neutron energy 
spectrum, power and breeding distribution, temperature coefficients of 
reactivity) of the \gls{MSBR} at startup and equilibrium state. A tertiary goal 
has been to demonstrate benefits of continuous fission products removal for 
thermal \gls{MSR} design.

Toward these goals, a full-core high-fidelity benchmark model of the \gls{MSBR} 
was implemented in SERPENT 2. The purpose of the full-core model instead of 
the simplified single-cell model \cite{betzler_molten_2017, 
rykhlevskii_online_2017, betzler_fuel_2018} was to precisely describe the 
two-region \gls{MSBR} concept design sufficiently to accurately represent 
breeding in the outer core zone. When running depletion calculations, the most 
important fission products and $^{233}$Pa are removed and fertile/fissile 
materials are added to the fuel salt every 3 days.  Meanwhile, the removal 
interval for the rare earths, volatile fluorides, and seminoble metals was more 
than month which causes effective multiplication factor fluctuation.

\subsection{Equilibrium state search}
The results of this study indicate that the effective multiplication factor 
slowly decreases from 1.075 and reaches 1.02 at equilibrium after approximately 
6 years of operation. At the same time, the concentration of $^{233}$U, $^{232}$Th, 
$^{233}$Pa, $^{232}$Pa stabilizes after approximately 2500 days of operation. 
Particularly, $^{233}$U number density equilibrates\footnote{fluctuates less 
than 0.8\%} after 16 years of operation. Consequently, the core reaches the quasi-equilibrium state 
after 16 years of operation. On the other hand, a wide diversity of nuclides, 
including fissile isotopes (e.g. $^{233}$U, $^{239}$Pu) and non-fissile strong 
absorbers (e.g. $^{234}$U), continue accumulating in the core. The current work results 
show that a true equilibrium composition cannot exist but the balance
between strong absorber accumulation and new fissile material 
production can be achieved to keep the reactor critical.

\subsection{Spectral shift}
We also found that the neutron energy spectrum grows harder as equilibrium 
approaches because significant heavy fission products accumulate in 
the \gls{MSBR} core. Moreover, the neutron energy spectrum in the central core 
region is much softer than in the outer core region due to lower 
moderator-to-fuel ratio in the outer zone, and this distribution remains stable 
during reactor operation. Finally, the epithermal or thermal spectrum is needed 
to effectively breed $^{233}$U from $^{232}$Th because radiative capture cross 
section of thorium-232 monotonically decreases from $10^{-10}$ MeV to $10^{-5}$ 
MeV. A harder spectrum in the outer core region tends to significantly increase 
resonance absorption in thorium and decrease the absorptions in fissile and 
structural materials.

The spatial power distribution in the \gls{MSBR} shows that 98\% of the fission 
power is generated in central zone I, and neutron energy spectral shift did not 
cause any notable changes in a power distribution. The spatial distribution of 
neutron capture reaction rate for fertile $^{232}$Th, corresponding to breeding in 
the core, confirms that most of the breeding occurs in an outer, 
undermoderated, region of the \gls{MSBR} core. Finally, the average $^{232}$Th 
refill rate throughout 60 years of operation is approximately 2.40 kg/day or 
100 g/GWh$_e$.

We compared the safety parameters for the initial fuel loading and 
equilibrium compositions using the SERPENT 2 Monte Carlo code. 
The total temperature coefficient 
is large and negative at startup and equilibrium but the magnitude decreases 
throughout reactor operation from $-3.10$ to $-0.94$ pcm/K as the spectrum 
hardens. The moderator 
temperature coefficient is positive and also decreases during fuel depletion. 
From reactivity control system efficiency analysis, showed the safety rod integral 
worth decreases by approximately 16.2\% over 16 years of operation, while 
graphite rod integral worth remains constant. Summing up, neutron energy 
spectrum hardening during fuel salt depletion has an undesirable impact on 
\gls{MSBR} stability and controllability, and should be taken into 
consideration in further analysis of accident transient scenarios.

\subsection{Benefits of fission product removal}

For thermal spectrum \gls{MSBR} removal of volatile gases, noble metals, and 
rare earths from a fuel salt benefits core performance.
Moreover, immediate removal of volatile gases (e.g., xenon) and noble metals 
increased reactivity by approximately 7500 pcm over a 10-year 
timeframe. In contrast, the effect of relatively slower removal of rare earth 
elements (every 50 days cycle instead of 3 days) has less impact (5500 pcm) on 
the core reactivity after 10 years of operation. In sum, additional study 
is needed to establish neutronic  and economic tradeoffs of removing each element.

\subsection{Future work}
Continued SaltProc-SERPENT coupled simulation efforts could progress in a 
number of different directions. First optimization of reprocessing parameters (e.g. time step, feeding rate, 
protactinium removal rate) could establish the best fuel utilization, breeding 
ratio, or safety characteristics for various designs. This might be performed with a parameter sweeping 
outer loop which would change an input parameter by a small increment, run the 
simulation and analyze output to determine optimal configuration. Alternatively, 
the existing RAVEN optimization framework \cite{alfonsi_raven_2013} might be 
employed for such optimization studies.

Only the batch-wise online reprocessing approach has been treated in this 
work. However, the SERPENT 2 Monte Carlo code was recently extended for 
continuous online fuel reprocessing simulation \cite{aufiero_extended_2013}. 
This extension must be verified against existing SaltProc/SERPENT or 
ChemTriton/SCALE packages, and could be employed for simulating immediate removal of 
fission product gases (e.g., Xe, Kr) which have a strong negative impact on 
core lifetime and breeding efficiency. Finally, using the built-in SERPENT 2 
Monte Carlo code online reprocessing \& refueling material burnup routine would 
significantly speed up computer-intensive full-core depletion simulations.

Lastly, this work enables accident safety analysis such as multi-physics 
simulation of the \gls{MSBR} with the coupled 
neutronics/thermal-hydraulics code, Moltres \cite{lindsay_introduction_2018}. 
The existing full-core SERPENT 2 model and equilibrium fuel material 
composition would be employed to generate problem-oriented nuclear data 
libraries for further usage in accident transient analysis. The final goal of 
this effort is to develop a fast-running computational model which could study 
the dynamic behavior of generic \glspl{MSR}, performing detailed safety 
analysis and design optimization for a variety of reactor concepts.

### 2018-08-15 methods.tex

The tool structure and capabilities of SaltProc are similar to the ChemTriton tool 
developed in \gls{ORNL} for SCALE \cite{powers_new_2013}. SaltProc is coupled with Monte 
Carlo SERPENT 2
software which allows to simulate online reprocessing for irregular full-core 
geometry with high level of fidelity.  The primary function of SaltProc is to 
manage material streams while
SERPENT 2 performs most of the computationally heavy work, namely neutron 
transport and depletion
calculations. Saltproc is defined as a python class, where each material stream 
is defined as a isotopic atomic density
vector variable. This allows tracking of time-sensitive material streams such 
as the
$^{233}$Pa tank in the \gls{MSBR}. The user can define the reprocessing 
parameters, such as the reprocessing interval and removal efficiency.  In 
addition, SaltProc provides a set of functions for each stream: read and write 
isotopic data in/from database, separate out specific isotopes from stream with 
defined efficiency, feed in specific isotopes to stream, and maintain constant 
number density of specific nuclide in the core. These attributes and functions 
are crucial to simulate the operation of a complex, multi-zone, multi-fluid 
\gls{MSR} and are universal for myriad reactor systems.

Saltproc is currently in active development in Github (https://github.com/ 
arfc/saltproc), and has unit tests and Travis continuous integration for 
sustainable development. There is also documentation
generated through Sphinx document generator for ease of use. We plan to 
implement
user-defined reprocessing scheme definition, two-region \gls{MSR} modeling 
capabilities,
and decay modeling in future releases.

Figure~\ref{fig:saltproc_flow} illustrates the  online reprocessing simulation 
algorithm the coupled SaltProc and SERPENT 2. To perform a depletion step,
SaltProc reads an external SERPENT 2 template file which is defined by the 
user. This file contains input cards with parameters such as geometry,
moderator and construction materials isotopic composition, neutron population, 
criticality cycles, total heating power, and boundary conditions.
After the depletion calculation, SaltProc reads the depleted fuel composition 
file (\texttt{
.bumat1}) and stores the depleted
composition isotopic vector in an HDF5 database. SaltProc only stores and edits 
the isotopic composition of the fuel stream,
which makes SaltProc a flexible tool to model any geometry: an infinite medium, 
a unit cell, a multi-zone simplified assembly, or a full core.
This flexibiliity allows the user to perform simulations of varying fidelity 
and computational intensity.
\begin{figure}[ht!] % replace 't' with 'b' to \centering
  \includegraphics[width=0.75\textwidth]{saltproc_flowchart.pdf}
  \caption{Flow chart for the Saltproc python-based tools.}
  \label{fig:saltproc_flow}
\end{figure}
SaltProc can manage as many material streams as desired. It also may work with 
multiple depletion materials. At the end of a each depletion step, SaltProc 
reads the depleted compositions and tracks each material stream individually. 
Following this, it applies chemical separation functions to fuel stream 
vectors. These vectors then form a matrix (isotopics x timesteps) which 
SaltProc stores in an HDF5 database and prints into the SERPENT 2 composition 
file for the next depletion calculation.

@@

In addition, SaltProc is able to define time-dependent material feed and 
removal rates to investigate the their impacts. These rates need not be 
constant in SaltProc. They can be defined as piecewise functions or set to 
respond conditions in the core. For instance, SaltProc might increase the 
fissile material feeding rate if the effective multiplication factor, 
$k_{eff}$, falls below a specific limit (e.g., 1.002).
These capabilities allow SaltProc to analyze fuel cycle of a generic 
liquid-fueled \gls{MSR}. In summary, the development approach of SaltProc 
focused on producing a generic, flexible and expandable tool to give the 
SERPENT 2 Monte Carlo code the ability to conduct advanced in-reactor fuel 
cycle analysis as well as simulate a myriad of online refueling and fuel 
reprocessing systems.

>>>>

The tool structure and capabilities of SaltProc are similar to the ChemTriton tool 
developed in \gls{ORNL} for SCALE \cite{powers_new_2013}. SaltProc is coupled 
with the Monte Carlo SERPENT 2
software to simulate online reprocessing for irregular full-core 
geometry with high fidelity.  The primary function of SaltProc is to 
manage material streams while
SERPENT 2 performs the computationally heavy work, namely neutron 
transport and depletion
calculations. Saltproc is defined as a python class, where each material stream 
is defined as a isotopic atomic density
vector variable. This allows tracking of time-sensitive material streams such 
as the
$^{233}$Pa tank in the \gls{MSBR}. The user can define the reprocessing 
parameters, such as the reprocessing interval and removal efficiency.  In 
addition, SaltProc provides a set of functions for each stream: read and write 
isotopic data in/from database, separate out specific isotopes from stream with 
defined efficiency, feed in specific isotopes to stream, and maintain constant 
number density of specific nuclide in the core. These attributes and functions 
are crucial to simulating the operation of a complex, multi-zone, multi-fluid 
\gls{MSR} and sufficiently general to represent myriad reactor systems.

SaltProc, currently in active development in Github (https://github.com/ 
arfc/saltproc), leverages unit tests and continuous integration for 
sustainable development. There is also documentation
generated through Sphinx document generator for ease of use. In future 
releases, we plan to 
implement
support for entirely user-customized reprocessing strategies, two-region \gls{MSR} modeling 
capabilities,
and decay modeling.

Figure~\ref{fig:saltproc_flow} illustrates the  online reprocessing simulation 
algorithm coupling SaltProc and SERPENT 2. To perform a depletion step, 
SaltProc reads a user-defined SERPENT 2 template file. This file contains input 
cards with parameters such as geometry, material, isotopic composition, neutron 
population, criticality cycles, total heating power, and boundary conditions.  
After the depletion calculation, SaltProc reads the depleted fuel composition 
file (\texttt{ .bumat1}) and stores the depleted composition isotopic vector in 
an HDF5 database.

SaltProc only stores and edits the isotopic composition of 
the fuel stream, which makes SaltProc a flexible tool to model any geometry: an 
infinite medium, a unit cell, a multi-zone simplified assembly, or a full core.
This flexibiliity allows the user to perform simulations of varying fidelity 
and computational intensity.

\begin{figure}[ht!] % replace 't' with 'b' to \centering
  \includegraphics[width=0.75\textwidth]{saltproc_flowchart.pdf}
  \caption{Flow chart for the Saltproc python package.}
  \label{fig:saltproc_flow}
\end{figure}
SaltProc can manage as many material streams as desired. It also may work with 
multiple depletion materials. At the end of a each depletion step, SaltProc 
reads the depleted compositions and tracks each material stream individually. 
Following this, it applies chemical separation functions to fuel stream 
vectors. These vectors then form a matrix (isotopics x timesteps) which 
SaltProc stores in an HDF5 database and prints into the SERPENT 2 composition 
file for the next depletion calculation.

@@

In addition, SaltProc is able to define time-dependent material feed and 
removal rates to investigate their impacts. These rates need not be 
constant in SaltProc. They can be defined as piecewise functions or set to 
respond to conditions in the core. For instance, SaltProc might increase the 
fissile material feeding rate if the effective multiplication factor, 
$k_{eff}$, falls below a specific limit (e.g., 1.002).
These capabilities allow SaltProc to analyze fuel cycle of a generic 
liquid-fueled \gls{MSR}. In summary, the development approach of SaltProc 
focused on producing a generic, flexible and expandable tool to give the 
SERPENT 2 Monte Carlo code the ability to conduct advanced in-reactor fuel 
cycle analysis as well as simulate a myriad of online refueling and fuel 
reprocessing systems.

### 2018-08-15 results.tex

The effective multiplication factor of the core reduces gradually over 
operation time because the fissile material ($^{233}$U) continuously depletes 
from the fuel salt due to fission and, at the same time, fission products 
accumulates in the fuel salt.	Eventually, without fission products removal, 
the reactivity decreases to the subcritical state after approximately 500 and 
1300 days of operation for cases with no removal and volatile gases \& noble 
metals removal, respectively. The time when the simulated core reaches 
subcriticality ($k_{eff}<$1.0) for full-core model) called the core lifetime. 
Therefore, removing fission products provides with significant neutronic 
benefit because allows achieve longer core lifetime.
\begin{figure}[ht!] % replace 't' with 'b' to force it to \centering
  \includegraphics[width=\textwidth]{keff_rem_cases.png} \caption{Calculated 
  effective multiplication factor for full-core \gls{MSBR} model with removal 
  of different fission product groups over 10 years of operation. The 
  confidence interval $\pm\sigma	$ is shaded.}
  \label{fig:fp_removal}
\end{figure}

>>>>

The effective multiplication factor of the core reduces gradually over 
operation time because the fissile material ($^{233}$U) continuously depletes 
from the fuel salt due to fission and, at the same time, fission products 
accumulates in the fuel salt.	Eventually, without fission products removal, 
the reactivity decreases to the subcritical state after approximately 500 and 
1300 days of operation for cases with no removal and volatile gases \& noble 
metals removal, respectively. The time when the simulated core reaches 
subcriticality ($k_{eff}<$1.0) for full-core model) called the core lifetime. 
Therefore, removing fission products provides with significant neutronic 
benefit because allows achieve longer core lifetime.
\begin{figure}[ht!] % replace 't' with 'b' to force it to \centering
  \includegraphics[width=\textwidth]{keff_rem_cases.png} \caption{Calculated 
  effective multiplication factor for full-core \gls{MSBR} model with removal 
  of various fission product groups over 10 years of operation. The 
  confidence interval $\pm\sigma	$ is shaded.}
  \label{fig:fp_removal}
\end{figure}
